Efficient pedigree recording for fast population genetics simulation

### 2017-11-17 forwards_paper.tex
Add most overview comments

\include{macros}

\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}

\usepackage{color}
\newcommand{\krt}[1]{{\em \color{green} #1}}
\newcommand{\plr}[1]{{\em \color{blue} #1}}
\newcommand{\jk}[1]{{\em \color{red} #1}}

@@

\begin{abstract}
    The computational burden of tracking individual genomes 
    in a population genetics simulation can be substantial.
    In this note we describe how to both (a) dramatically reduce this burden and
    (b) efficiently record the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
    on which neutral mutations can be quickly placed afterwards --
    the \emph{tree sequence} introduced in the software package \msprime.
    We implement the method in two popular fowards-time simulation frameworks,
    and show it leads to order-of-magnitude speedups.
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards,
without loss of generality.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we show how to use algorithmic tools and data structures developed for the 
coalescent simulator \msprime{}
to efficiently record, and later process, this history.

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths1997ancestral},
which has been the subject of substantial study
under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.

\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
Forwards-time population genetics simulations are growing in importance,
as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
For instance, \citet{harris2016genetic} used SLiM \citep{haller2017flexible} to simulate tens of thousands of human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
This is roughly state-of-the-art 
-- the fastest simulation frameworks such as SLiM and fwdpp \citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations --
but still substantially short of whole genomes for most biological population sizes.
There has been some progress at parallelization \citep{lawrie2017accelerating},
which has substantial promise.
The general idea of storing genealogical information to speed up simulations is not new --
it was entirely implemented in AnA-FiTS \citep{aberer2013rapid}, for instance --
but this program did not implement the critical step of discarding irrelevant genealogical information.
A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.

@@

Below, we first describe the conceptual and algorithmic foundations for the method:
(a) the new \emph{tables} format for storing tree sequences;
(b) an algorithm to output tree sequence information on the fly
    from a forwards-time simulation;
and (c) a method to \emph{simplify} a tree sequence, i.e., remove redundant information.
This is followed by benchmarking comparisons
with real, forwards-time simulations.

%%%%%%
\subsection*{Recording a tree sequence}

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
with overlapping generations that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random 
(and each such instance represents and independent draw).

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals,
with a probability of death per generation $\delta$ and a chromosome of length
$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
describing the recorded edges.
}

@@

\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x, L, P_b, n) \right\}$
    and $n \leftarrow n + 1$.
}

@@

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It has 10 node records (one for each individual)
        and 14 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis --
        for instance, $C$ and $D$ were parents to $F$,
        who inherited the left 70\% of the chromosome from $D$ and the remainder from $C$.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red.
        % which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

The tree sequences produced by forwards simulations, 
record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of the current population.
Simplification is essential to reduce this 
to a manageable quantity that still contains all
the information that we are interested in. 
Simplification is also useful if we have a
large tree sequence representing a large dataset and we wish to extract the
information relevant to a subset of the samples.

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
Conceptually, this works by 
(a) beginning by painting the chromosome each sample a distinct color;
(b) moving back through history, 
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color; 
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
The process is depicted in Figure~\ref{fig:simplify_state}.

@@

More concretely,
the algorithm works by moving back through time,
processing each parent in the input tree sequence in chronological order.
The main state of the algorithm at each point in time is a set of ancestral lineages,
and each lineage is a collection of ancestral segments.
An ancestral segment $(\ell, r, u)$ is found in a lineage
if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
% These segments are stored in a collection of linked lists,
% one list of segments for each ancestral lineage present at that time.
(These lineages are the ``colors'' above.)
We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.
Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence..

\paragraph{Sequential simplification}
At any point the simulation scheme above, genealogical history is recorded
in a tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime.

%%%%%%%%%%%%
\subsection*{Estimates of run-time complexity}

Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of 
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that 
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.

@@

Comparison of simulation with/without msprime, using simuPOP
or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).

@@

A final note:
in preparing this manuscript, 
we debated a number of possible terms for the embellished pedigree,
i.e., the ``pedigree with ancestral recombination information''.
Current etymologica consensus \citep{liberman2014little} has
``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
(whose branching pattern resembes the bifurcation of a single parent-offspring relationship).
An analogous term for the embellished pedigree might be \emph{nedigree},
from ``nid de grue'',
as the nest of a crange is a large jumble of (forking) branches.
We thought it unwise to use this term throughout the manuscript,
but perhaps it will prove useful elsewhere.

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}

\usepackage{color}
\newcommand{\krt}[1]{{\em \color{green} #1}}
\newcommand{\plr}[1]{{\em \color{blue} #1}}
\newcommand{\jk}[1]{{\em \color{red} #1}}
\newcommand{\jda}[1]{{\em \color{cyan} #1}}

@@

\begin{abstract}
    The computational burden of tracking individual genomes 
    in a population genetics simulation can be substantial.
    In this note we describe how to both (a) dramatically reduce this burden and
    (b) efficiently record the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
    on which neutral mutations can be quickly placed afterwards ---
    the \emph{tree sequence} introduced in the software package \msprime.
    We implement the method in two popular fowards-time simulation frameworks,
    and show it leads to order-of-magnitude speedups.
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards.
\jda{WLOG seems unclear/ not needed}
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, describe algorithmic tools and data structures developed for the 
coalescent simulator \msprime{}
to efficiently record, and later process, this history.

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph} ---
the {ARG} \citep{griffiths1997ancestral},
where nodes are common ancestor or recombination events ---
which has been the subject of substantial study
under the assumptions of coalescent theory \citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
\jda{would be nice to say, concisely the difference}

\jda{Is this the right place for this paragraph? should it maybe be integrated
	with graf beginning 'Modern computing' above?}
\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
Forwards-time population genetics simulations are growing in importance,
as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
For instance, \citet{harris2016genetic} used SLiM \citep{haller2017flexible} to simulate tens of thousands of human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
This is roughly state-of-the-art 
-- the fastest simulation frameworks such as SLiM and fwdpp \citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations --
but still substantially short of whole genomes for most biological population sizes.
There has been some progress at parallelization \citep{lawrie2017accelerating},
which has substantial promise.
\jda{this is vague, promise for what?}
The general idea of storing genealogical information to speed up simulations is not new --
it was entirely implemented in AnA-FiTS \citep{aberer2013rapid}, for instance --
but this program did not implement the critical step of discarding irrelevant genealogical information.
A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.

@@

Below, we first describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for storing tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) a method to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a method that combines (b) and (d) to allows efficiently record simplified a tree sequence,
and present analyses of its run time complexity.
This is followed by benchmarking comparisons between two implementations of our method,
each using previously published forwards-time simulation libraries.

%%%%%%
\subsection*{Storing a tree sequence}

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
(but with overlapping generations) that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
\jda{In the below we are using the integer vector encoding of a tree as in
	\msprime{} should say so and reference it as the role of node ids is
	otherwise more confusing than it need be}
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random 
(and each such instance represents and independent draw).

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population $P$ of $N$ haploid individuals,
with a probability of death per generation $\delta$ and a chromosome of length
$L$ for $T$ generations. On termination, there will be $n$ nodes with times recorded
in the vector $\tau$, and $E$ will contain a set of $(\ell, r, p, c)$ tuples
describing the recorded edges.
}

@@

\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, n - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, n - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

\algstep{W7.}{Record edges}{Set $E \leftarrow E \cup \left\{ (0, x, P_a, n), (x,
		L, P_b, n) \right\}$.
}

\algstep{W8.}{Increment the current node.}{
$n \leftarrow n + 1$.
}

@@

\begin{figure}
    \begin{center}
        \includegraphics{method_diagram}
    \end{center}
    \caption{
        A simple example of the simplify method.
        \textbf{Top:} the diagram on the left relates ten haploid individuals to each other.
        It has 10 node records (one for each individual)
        and 14 edge records (one for each distinctly inherited segment).
        Blue numbers denote crossing over locations in each meiosis --
        for instance, $C$ and $D$ were parents to $F$,
        who inherited the left 70\% of the chromosome from $D$ and the remainder from $C$.
        The individuals $B$, $C$, and $D$ inherit clonally from $A$.
        \textbf{Center:} the five distinct trees relating all individuals to each other
        found across the chromosome (blue numbers denote locations on the chromosome).
        Labels after simplification are shown in red.
        \textbf{Bottom:} tables recording the tree sequence after simplification
        with nodes $I$ and $J$ as samples.
        The mapping from labels in the forwards time simulation to nodes in the tree sequence
        is shown in red.
        % which allows additional records to be added as the simulation progresses.
        \label{fig:method_diagram}
    }
\end{figure}

The tree sequences produced by forwards simulations 
record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of the current population.
Simplification is essential to reduce this 
to a manageable quantity that still contains all
the information that we are interested in. 
Simplification is also useful if we have a
large tree sequence representing a large dataset and we wish to extract the
information relevant to a subset of the samples.

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
Conceptually, this works by 
(a) beginning by painting the chromosome each sample a distinct color;
(b) moving back through history, 
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color; 
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
The process is depicted in Figure~\ref{fig:simplify_state}.
\jda{Need refs to Figure~\ref{fig:method_diagram} in the text}

@@

More concretely,
the algorithm works by moving back through time,
processing each parent in the input tree sequence in chronological order.
The main state of the algorithm at each point in time is a set of ancestral lineages,
and each lineage is a collection of ancestral segments.
An ancestral segment $(\ell, r, u)$ is found in a lineage
if the output node $u$ inherits the genomic interval $[\ell, r)$ from that lineage.
% These segments are stored in a collection of linked lists,
% one list of segments for each ancestral lineage present at that time.
(These lineages are the ``colors'' above.)
We also maintain a map $A$ such that $A_j$ is segment chain for node $j$ in
the input tree sequence.
Crucially, the time required to run the algorithm is linear in the number of edges of the input tree sequence.

\subsection*{Forward simulation and sequential simplification to record simplifed tree sequences}

At any point in Algorithm~\algref{W}, genealogical history is recorded
in an unsimplified tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime. 
This process of forward simulation coupled with sequential simplification is our
method

\jda{I didn't write down the full thing, but mightn't things be clearer if we
	have a statement of the combined method?}
\begin{taocpalg}{F}{Forwards-time simplifed tree sequence}{
Simulates a population starting with $N$ diploid individuals on a chromosome of length $L$ for $T$
generations.
}

\algstep{F0.}{Initialize nodes.}{
Generate initial haploid nodes $\mathcal N$:
1.) using coalsecent simulation with \msprime{} for $2N$ initial haploid samples $S$ on
a chromosome of length $L$ and taking $\mathcal N = \{\texttt{id}(u) : u \in S \}$; or
2.) simply defining $\mathcal N = \{1\dots 2N\}$ as the node ids.
Define a mapping from  $\mathcal M: \{1\dots 2N\} \to \mathcal N$. In the second
case $\mathcal M$ is the identity.
}

\algstep{F1.}{Initialize edges.}{
Define the initial edges $E_0$: 1) either from the coalescent simulation's final
state, or 2) as \dots
}

\algstep{F2.}{Record trees till time $T - \delta$.}{ 
Simulate, using Algorithm~\algref{W} or another algorithm, $2N$ haploid
individuals using $\mathcal M$ to renumber the nodes, producing nodes $n$ and
edgesets $E$.
}

\algstep{F3.}{Simplification over $\delta$.}{
Select final samples $S'$ and using the simplification Algorithm~\algref{S},
simplify $n$ and $E$ to produce a simplified tree sequence.
}

\algstep{F4.}{Continue?}{
	Stop if $T - \delta  = 0$, else continue.
}
\end{taocpalg}

%%%%%%%%%%%%
\subsubsection*{Estimates of run-time complexity}

\jda{I'd like to outline what we're doing here with reference to alg F}
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals 
using Algorithm~\algref{W}
for $T$ generations,
with the probability of death, $\delta$, set to 1 for simplicity.
Since there is exactly one crossing over every generation,
this produces tables of 
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
so that $T \sim 20N$, say.
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
This tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
three edges end and three new edges begin (except for changes affecting the root; see \citet{kelleher2016efficient}).
Coalescent theory tells us that 
the expected total length of edges in a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per generation.
Not all such recombinations actually change the tree topology,
and three times this gives us an upper bound on the expected number of edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $160 N^2$ items to $8N(4\log(2N) + 3)$;
at $N=10^4$, a factor of 4,700.

@@

Comparison of simulation with/without msprime, using \simupop{}
or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).

@@

A final note:
in preparing this manuscript, 
we debated a number of possible terms for the embellished pedigree,
i.e., the ``pedigree with ancestral recombination information''.
Current etymologica consensus \citep{liberman2014little} has
``pedigree'' derived from the french ``pied de grue'' for the foot of a crane
(whose branching pattern resembes the bifurcation of a single parent-offspring relationship).
An analogous term for the embellished pedigree might be \emph{nedigree},
from ``nid de grue'',
as the nest of a crane is a large jumble of (forking) branches.
We thought it unwise to use this term throughout the manuscript,
but perhaps it will prove useful elsewhere.
